---
title: "Possible COVID-19 Infection Trajectories in South Africa"
author: "Rendani Mbuvha, Tshilidzi Marwala"
date: "3/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This notebook aims to provide some initial high-level forecasting of the progression of COVID-19 in South Africa by drawing lessons from progression in other countries. **At the moment given the sparse data, the aim is not forecasting precision but scenario mapping.**

Essentially, this notebook tries to use simple curve fitting methods to answer the questions: What would happen in the next 60 days if South Africa's COVID-19 infections followed the infection trajectory of:

* China
* Italy
* South Korea
* Iran
* The "World" (global average) 

Another question we try to answer is what if we use Facebook's [Prophet](https://facebook.github.io/prophet/) model to forecast South Africa's trajectory.  

The key takeaway is that the Chinese trajectory is the best possible case --Italy, Iran and (even) South Korea would be nightmare scenarios for us.

**Note to reader** - All models are fitted on cases per million and then scaled up to SA's population at the end.



##Loading the relevent libraries
```{r echo = FALSE}
library(prophet)
library(ggplot2)
library(reshape2)
```

##Reading in the Data
The data included in this analysis is upto and including the 24th of March 2020



```{r}
total_cases = read.csv("total_cases.csv")
total_cases_per_mil = read.csv("total_cases_per_mil.csv")
total_cases_per_mil_adj = read.csv("total_cases_per_mil_test_adj.csv")
total_cases_per_mil_adj_expand = read.csv("total_cases_per_mil_test_adj_expand.csv")
population = read.csv("poputlation.csv")
index = seq(1,85)
total_cases_per_mil_adj =cbind(total_cases_per_mil_adj,index)
```
##Initial High-level Plots
```{r}
total_cases_per_mil_adj_expand$date = as.Date(total_cases_per_mil_adj_expand$date, format = "%d/%m/%Y")
ggplot(total_cases_per_mil_adj_expand[1:680,], aes(x = date, y = Confirmed_Cases)) + 
  geom_line(aes(color = Country), size = 1) +
    xlab("Date") + ylab("Cases Per Million")+
  ggtitle("Confirmed Cases Per Million by Country")+
  theme_minimal()
```

Observations

* Italy has huge problems
* China and South Korea have managed to flatten the curve
* Iran seems to be working hard on this
* UK should be worried but... they test more than the average!
* SA(and the 'World') looks okay for now -- We did adjust for lower testing in SA(this seems to increase cases by about 2% *i.e if SA was testing similar to global average*).

```{r include = FALSE}
sa_data = total_cases_per_mil_adj[67:85,c("South.Africa","date")]
sk_data= total_cases_per_mil_adj[21:85,c("South.Korea","date")]
iran_data=total_cases_per_mil[52:85,c("Iran","date")]
china_data=total_cases_per_mil_adj[1:85,c("China","date")]
italy_data=total_cases_per_mil[32:85,c("Italy","date")]
world_data=total_cases_per_mil_adj[1:85,c("World","date")]


names(sa_data)[2] = "ds"
names(sk_data)[2] = "ds"
names(iran_data)[2] = "ds"
names(italy_data)[2] = "ds"
names(world_data)[2] = "ds"
names(china_data)[2] =  "ds"

names(sa_data)[1] = "y"
names(sk_data)[1] = "y"
names(iran_data)[1] = "y"
names(italy_data)[1] = "y"
names(world_data)[1] = "y"
names(china_data)[1] =  "y"

sa_data$ds = as.Date(sa_data$ds , format = "%d/%m/%Y")
sk_data$ds = as.Date(sk_data$ds, format = "%d/%m/%Y")
iran_data$ds = as.Date(iran_data$ds, format = "%d/%m/%Y")
italy_data$ds = as.Date(italy_data$ds, format = "%d/%m/%Y")
world_data$ds = as.Date(world_data$ds, format = "%d/%m/%Y")
china_data$ds = as.Date(china_data$ds, format = "%d/%m/%Y")

sa_data_2 = sa_data
sa_data_2$y = log(sa_data_2$)
```
##Facebook Prophet time series forecasting model on South African Data

These model seems to require more data or perhaps a saturation point to be predefined. Perhaps we must return next month(more data)??

```{r echo = FALSE}

sa_model = prophet(sa_data,n.changepoints=3)
```

Create future periods for South Africa i.e next 60 days
```{r echo = FALSE}
sa_future = make_future_dataframe(sa_model, periods = 60)

```
Make Predictions for South Africa
```{r}
sa_forecast = predict(sa_model , sa_future)
tail(sa_forecast[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])

```

###Plot South Africa's Forecasts(still at per million level)

```{r}
plot(sa_model, sa_forecast)
```

Look at some of the forecast components

```{r}
prophet_plot_components(sa_model
                        , sa_forecast)
```
```{r}
sa_future$time <- seq.int(nrow(sa_future))
sa_data$time <- seq.int(nrow(sa_data))
sk_data$time <- seq.int(nrow(sk_data))
iran_data$time <- seq.int(nrow(iran_data))
china_data$time <- seq.int(nrow(china_data))
italy_data$time <- seq.int(nrow(italy_data))
world_data$time <- seq.int(nrow(world_data))
```
##Curve Fitting
###Second-order polynomial
```{r}

world_2o= lm(y~poly(time-1,2,raw=TRUE),data = world_data) 
china_2o= lm(y~poly(time-1,2,raw=TRUE), data = china_data) 
italy_2o= lm(y~poly(time-1,2,raw=TRUE), data = italy_data) 
iran_2o= lm(y~poly(time-1,2,raw=TRUE), data = iran_data) 
south_korea_2o= lm(y~poly(time-1,2,raw=TRUE), data = sk_data) 
south_africa_2o= lm(y~poly(time-1,2,raw=TRUE), data = sa_data) 
```
###Goodness of Fit
with an R_squared of 0.97 this model is not bad i.e caputure 97% of variance in the current data
```{r}
plot(sa_data$time,sa_data$y,pch=19)
lines(sa_data$time,predict(south_africa_2o,sa_data),col="red")
summary(south_africa_2o)
```
###Making Predictions Based On Other Countries(second-order polynomial)

```{r include = FALSE}
world_2o_predictions = as.data.frame((predict(world_2o,sa_future))*population[5,2]/1000000)
world_2o_predictions[1:19,]=sa_data$y*population[5,2]/1000000
china_2o_predictions= as.data.frame((predict(china_2o,sa_future))*population[5,2]/1000000)
china_2o_predictions[1:19,]=sa_data$y*population[5,2]/1000000
italy_2o_predictions= as.data.frame((predict(italy_2o,sa_future))*population[5,2]/1000000)
italy_2o_predictions[1:19,]=sa_data$y*population[5,2]/1000000
south_korea_2o_predictions=as.data.frame((predict(south_africa_2o,sa_future))*population[5,2]/1000000)
south_korea_2o_predictions[1:19,]=sa_data$y*population[5,2]/1000000
south_africa_2o_predictions=as.data.frame((predict(south_africa_2o,sa_future))*population[5,2]/1000000)
iran_2o_predictions = as.data.frame((predict(iran_2o,sa_future))*population[5,2]/1000000)
iran_2o_predictions[1:19,]=sa_data$y*population[5,2]/1000000
```
```{r include = FALSE}
colnames(iran_2o_predictions)="forecast"
colnames(south_africa_2o_predictions)="forecast"
colnames(south_korea_2o_predictions)="forecast"
colnames(italy_2o_predictions) ="forecast"
colnames(china_2o_predictions)="forecast"
colnames(world_2o_predictions)="forecast"
Trajectory =as.data.frame(c(rep("World",79),rep("China",79),rep("Italy",79),rep("South Korea",79),rep("Iran",79),rep("Current ZA",79)))
o2_predictions = rbind(as.data.frame(world_2o_predictions),as.data.frame(china_2o_predictions),as.data.frame(italy_2o_predictions),as.data.frame(south_korea_2o_predictions),as.data.frame(iran_2o_predictions),as.data.frame(south_africa_2o_predictions))
o2_pred_data_frame =cbind(rep(sa_future$ds,6),o2_predictions,Trajectory)
colnames(o2_pred_data_frame)=c("date","forecast","Trajectory")
```
```{r}
ggplot(o2_pred_data_frame, aes(x = date, y = forecast)) + 
  geom_line(aes(color = Trajectory), size = 1) +
    xlab("Date") + ylab("Case Forecast")+
    ggtitle("What if SA followed Country X:2nd Order Polynomial")
  #+geom_hline(yintercept=120000, linetype="dashed", color = "red")
 #+annotate("text", x = o2_pred_data_frame$date[30], y = 120010, label = "Total National Bed Capacity")

```

##Italy and Iran skew things lets not show them ...


```{r include = FALSE}
#colnames(iran_2o_predictions)="forecast"
colnames(south_africa_2o_predictions)="forecast"
colnames(south_korea_2o_predictions)="forecast"
#colnames(italy_2o_predictions) ="forecast"
colnames(china_2o_predictions)="forecast"
colnames(world_2o_predictions)="forecast"
Trajectory =as.data.frame(c(rep("World",79),rep("China",79),rep("Current ZA",79)))
o2_predictions = rbind(as.data.frame(world_2o_predictions),as.data.frame(china_2o_predictions),as.data.frame(south_africa_2o_predictions))
o2_pred_data_frame =cbind(rep(sa_future$ds,3),o2_predictions,Trajectory)
colnames(o2_pred_data_frame)=c("date","forecast","Trajectory")
```
```{r}
ggplot(o2_pred_data_frame, aes(x = date, y = forecast)) + 
  geom_line(aes(color = Trajectory), size = 1) +
    xlab("Date") + ylab("Case Forecast")+
  ggtitle("What if SA followed Country X:2nd Order Polynomial")+
  theme_minimal()
```









